---
title: "Open and plot shapefiles"
author: "Joseph Stachelek"
date: "October 22, 2015"
output: html_document
---

##About
This post explains the how to open and plot point, line, and polygon shapefiles in R.

###Goals / Objectives
After completing this activity, you will:

 * Understand the concept of point, line, and polygon shapefiles
 * Know how to open shapefiles (vector data)
 * Understand the components of a _spatial object_ (in R)
 
###Tools To Install

```{r Load Libraries}
library(rgdal, quietly = TRUE)
library(raster)
```

###Recommended Pre-Lesson Reading

#About Vector Data
##Shapefiles: Points, Lines, and Polygons
Vector geospatial data can be expressed as collections of points, lines, and polygons. This data is most commonly held in _shapefiles_ which store information about the geometry of the geospatial data as well as _attributes_ that describe the data. More about shapefiles can found on [Wikipedia](https://en.wikipedia.org/wiki/Shapefile) (add link to Metadata lesson?).

##Importing shapefiles

We are going to use the `R` function `readOGR()` to import shapefiles stored on disk. These shapefiles include a polygon shapefile representing the boundaries of our study area, a line shapefile representing roads and streams, and a point shapefile representing the location of a NEON measurement tower.

```{r Import Shapefile}
prepath <- "NEON-DC-DataLesson-Hackathon/_posts"
#prepath <- ".."

#Import a polygon shapefile 

squarePlot <- readOGR(file.path(prepath, "/data/boundaryFiles/HARV/"),"HarClip_UTMZ18")

#Import a line shapefile

lines <- readOGR(file.path(prepath, "/data/boundaryFiles/HARV/"),"HARV_roadStream")

#Import a point shapefile 

point <- readOGR(file.path(prepath, "/data/boundaryFiles/HARV/"),"HARVtower_UTM18N")
  
```

We can view an attribute summary of each shapefile by simplying entering the name of the `R` object in the console. Note that the output includes the _class_, the number of _features_, the _extent_, the _coordinate reference system_ (crs), and the _attributes_ of the`R` object.  

```{r View Attribute Summary}
#view attributes
squarePlot
lines
point
```

##Plotting shapefiles

The `plot()` function can be used for basic plotting of these spatial objects. Note that we use the `add = TRUE` argument to _overplot_ the lines and point shapefile onto the polygon shapefile. 

```{r Plotting}
#look at the shapefiles
plot(squarePlot, col = "purple")
plot(lines, add = TRUE)
plot(point, add  = TRUE, pch = 19, col = "red")
```

We can use the `col` argument to the `plot()` function to color our lines according to the _TYPE_ attribute. The following code chunk extracts a vector of this attribute, creates a character color vector of the same length, and changes the positions of this color vector that correspond to _footpath_.

```{r Color lines by attribute}
#color lines by TYPE attribute
type <- slot(lines, "data")$TYPE
col <- rep("black", length = length(type))
col[type == "footpath"] <- "red"

plot(lines, col = col, add = TRUE)
```
