---
layout: post
title: "Lesson 01: Shapefile Metadata & Attributes in R"
date:   2015-10-25
authors: [Joseph Stachelek, Leah Wasser]
contributors: [Megan A. Jones, Sarah Newman]
dateCreated:  2015-10-23
lastModified: `r format(Sys.time(), "%Y-%m-%d")`
packagesLibraries: [rgdal, raster]
category: 
mainTag: spatial-vector-series
tags: [vector-data]
description: "This lesson provides an overview of how to locate and query
shapefile attributes as well as subset shapefiles by specific attribute values
in R."
code1: 01-shapefile-attributes.R
image:
  feature: NEONCarpentryHeader_2.png
  credit: A collaboration between the National Ecological Observatory Network (NEON) and Data Carpentry
  creditlink: http://www.neoninc.org
permalink: /R/shapefile-attributes-in-R/
comments: false
---

{% include _toc.html %}

##About
This lesson explains what shapefile attributes are. It also covers how to
work with shapefile attributes in `R`. It covers how to identify and query 
shapefile attributes as well as subset shapefiles by specific attribute values. 
Finally, we will review how to plot a shapefile according to a set of attribute 
values.

**R Skill Level:** Intermediate - you've got the basics of `R` down.

<div id="objectives" markdown="1">

##Goals / Objectives
After completing this activity, you will:

 * Be able to query shapefile attributes.
 * Be able to subset shapefiles using specific attribute values..
 * Know how to plot a shapefile, colored by unique attribute values.
 
##Things You'll Need To Complete This Lesson
You will need the most current version of `R` and, preferably, RStudio loaded on
your computer to complete this lesson.

##R Libraries to Install:

* **raster:** `install.packages("raster")`
* **rgdal:** `install.packages("rgdal")`
* **sp:** `install.packages("sp")`

<a href="{{ site.baseurl }}/R/Packages-In-R/" target="_blank"> 
More on Packages in R - Adapted from Software Carpentry.</a>

##Data to Download
{% include/dataSubsets/_data_Site-Layout-Files-HARV.html %}

**Set Working Directory:** This lessons assumes that you have set your working 
directory to the location of the downloaded and unzipped data subset. 
[An overview of setting the working directory in `R` can be found here.]({{site.baseurl}}/R/Set-Working-Directory "R Working Directory Lesson") 

**Challenge Code:** NEON Data Skills tutorials often contain challenges that
reinforce learned skills. If available, the code for challenge solutions is
found in adownloadable `R` script available on the footer of each lesson page.

**Vector Lesson Series: ** This lesson is a part of a lesson series on vector 
data in `R`.

</div>

#Shapefile Metadata & Attributes
When we import a shapefile into `R`, the `readOGR()` function automatically
stores metadata and attributes associated with the file.

##Load the Data
To work with vector data in `R`, we can use the `rgdal` library. The `raster` 
package also allows us to explore metadata using similar commands for both
rasters and vectors. We will import shapefiles with data for a field site
boundary, roads within the field site, and the tower location.
If you completed the
[Vector Data in R - Open and Plot Shapefiles]({{site.baseurl}}R/open-shapefiles-in-R/ ) 
lesson, you can skip this code chunk.

```{r load-packages-data }
#load packages
#rgdal: for vector work; sp package should always load with rgdal. 
library(rgdal)  
#raster: for metadata/attributes- vectors or rasters
library (raster)   

#set working directory to data folder
#setwd("pathToDirHere")

#Import a polygon shapefile 
aoiBoundary_HARV <- readOGR("NEON-DS-Site-Layout-Files-HARV/HARV/", "HarClip_UTMZ18")

#Import a line shapefile
lines_HARV <- readOGR( "NEON-DS-Site-Layout-Files-HARV/HARV/", "HARV_roads")

#Import a point shapefile 
point_HARV <- readOGR("NEON-DS-Site-Layout-Files-HARV/HARV/", "HARVtower_UTM18N")

```

#Query Shapefile Metadata 
Shapefile metadata can include object `class`, a features count (`length`), the
object `extent`, and the object's coordinate reference system (`CRS`). 

```{r view-shapefile-metadata }
#view class
class(x = point_HARV)

# x= isn't actually needed; it just specifies which object
#view features count
length(point_HARV)

#view crs - note - this only works with the raster package loaded
crs(point_HARV)

#view extent- note - this only works with the raster package loaded
extent(point_HARV)

```

When we type in the full file name we see the same three types of metadata with
some additional attribute information.  

```{r view-shapefile-metadata2 }
point_HARV
```

<div id="challenge" markdown="1">
##Challenge: Metadata for Different Spatial Classes
Determine the `extent`, `CRS`, `length`, and `class` for the `lines_HARV` and 
`aoiBoundary_HARV` objects.  
Will all three objects (`lines_HARV`, `aoiBoundary_HARV`, & `point_HARV`) be
able to be mapped together? 
</div>
```{r challenge-code-metadata, results="hide", echo=FALSE}
#view class
class(lines_HARV)
class(aoiBoundary_HARV)

#view features count
length(lines_HARV)
length(aoiBoundary_HARV)

#view crs
crs(lines_HARV)
crs(aoiBoundary_HARV)

#view extent
extent(lines_HARV)
extent(aoiBoundary_HARV)

# All three objects have the same CRS and the extents overlap or at least are
# close so they should be able to appear on the same map. 
```

#Query Shapefile Attributes
Shapefiles often contain an associated database or spreadsheet of values called
`attributes` that describe the vector features in the shapefile. You can think
of this like a spreadsheet with rows and columns. Each column in the spreadsheet
is an individual `attribute` that describes an object (sometimes called
variables). Shapefile `attributes` include measurements that correspond to the
geometry of the shapefile features.

For example, the Roads shapefile (`lines_HARV` object) contains an attribute
called `TYPE`. Each line in the shapefile has an associated TYPE which describes
the type of road (road, footpath, boardwalk, etc).  

We can look at all of the associated data attributes by printing the contents of
the `data` slot with `@data`. 

```{r shapefile-attributes}
#just view the attributes of the data
(lines_HARV@data)
```

<i class="fa fa-star"></i> **Data Tip:** If you are comfortable using `slot()`,
it is another method to do the same thing.
{: .notice}

Using `@data` we can view a table with attribute names across the top and rows
for each line (or point/ polygon if using those) in the shapefile. The 0-12
printed in the first column is simply a `row.name`. We can test this by issuing
the following command to the `R` console: `row.names(aoiBoundary@data)`.

``` {r row-names}
#What are the row names?
row.names(lines_HARV@data)
```

If we want to select something about the attributes, we can do this by 
combining common functions (e.g. `names` or `head`) and then using `NAME@data`.

Let's give it a try.

```{r view-shapefile-attributes }
#view just the attribute names for the lines spatial object
names(lines_HARV@data)

#view just the top 5 attributes for the lines spatial object
head(lines_HARV@data)

```

Alternatively, we can find attributes by using the `attributes` function. Use
this with caution, if we have lots of objects in a shapefile this output will 
be very long. 

```{r view shapefile-attributes-2}
attributes(lines_HARV)

```

<div id="challenge" markdown="1">
##Challenge: Attributes for Different Spatial Classes
Explore the attributes associated with the `point_HARV` and `aoiBoundary_HARV` spatial objects. 

1. How many attributes do each have?
2. Who owns the site in the `point_HARV` data object?
3. Which of the following are NOT attributes of the `point` data object?
    A) Lat B) County C) Country
</div>
```{r challenge-code-attributes-classes, results="hide", echo=FALSE}
#1
names(point_HARV@data)  #14 attributes
names(aoiBoundary_HARV@data)  #1 attribute

#2
head(point_HARV@data)  #Harvard University, LTER

#3
point_HARV@data  # C Country
```

#Explore Values in One Attribute
We can also explore individual values stored within a particular attribute.
Again, comparing attributes to a spreadsheet or a `data.frame`, this is similar
to exploring values in a column. We can do this using the `$` and the name of
the attribute. 

```{r explore-attribute-values }
#view all attributes in the lines shapefile within the TYPE field
lines_HARV$TYPE

#view unique attributes in the lines shapefiles within the TYPE field
levels(lines_HARV$TYPE)

```

Levels is particularly useful if we have an object with lots of different lines,
polygons or points as it summarizes the differences for us.  

#Access Levels for Categorical Attributes
At times we may want to work with a smaller subset of all the data contained in
a shapefile. For example, in `lines_HARV` we've seen that there are different 
types of roads.  We might want to create a map were each of the different road
types is designated in a different color.  Or we might want to figure out a 
total distance of all the stone walls in the study site.  Both of these tasks 
can be accomplished by accessing the levels of this categorical attribute. 

##Subset Shapefiles
Using the `$` symbol, we were able to access the values associated with a 
particular `attribute` in a shapefile. We can use this syntax to select 
a subset of features from a spatial object in `R`. 

```{r Subsetting-shapefiles}
#view all attributes in the TYPE column
lines_HARV$TYPE

# select features that are of TYPE "footpath"
# could put this code into other function to only have that function work on
# "footpath" lines
lines_HARV[lines_HARV$TYPE == "footpath",]

#save an object with only footpath lines
footpath_HARV<-lines_HARV[lines_HARV$TYPE == "footpath"]
footpath_HARV
```

Our subsetting operation reduces the `features` count from 13 to 2. 

#Plot Vector Data in R
First, we need to plot all components: boundary, road, towers.  

```{r plot-multiple-shapefiles }
#Plot multiple shapefiles
plot(aoiBoundary_HARV, col = "lightgreen", 
     main="NEON Harvard Forest\nField Site")
plot(lines_HARV, add = TRUE)

plot(point_HARV, add  = TRUE, pch = 19, col = "purple")
```

Then we can modify different subsets of the data and plot them in different
colors.  To differentiate between the lines we need to add a figure legend.

We use the `col` argument in the `plot()` function to specify the color of
spatial objects. If type is an attribute of our data, we can use the syntax
`$TYPE` to tell `R` to plot the spatial objects according to their unique TYPE
values. 

The following code chunk extracts a vector of the attribute, creates 
a character color vector of the same length, and changes the positions 
of this color vector that correspond to `footpath`.

```{r Color-lines-by-attribute}
#color lines by TYPE attribute

#1st: create object that is the TYPE attribute 
# alternatively: change in the col() below but this object creates cleaner code)
lineType <- lines_HARV@data$TYPE

# Bonus: syntax if you prefer using slot(). 
# lineType <- slot(object = lines, name = "data")$TYPE

#2nd: set the color for all the lines 
col <- rep(x= "black", length = length(lineType))

#3rd: set the color you want for the TYPE level you want (changes it in the col
# vector)
col[lineType == "footpath"] <- "red"

#4th: plot. add=TRUE because we want this to map onto the previous plot 
plot(x=lines_HARV, col=col, add=TRUE)

#5th: Add legend so that we know what the colors represent. 
labels= c("footpath", "tower")
labCols=c("red", "purple")

# legend syntax: "bottom"= put legend in center bottom, bty="n" is no boarder
# around legend, cex=0.8 is the font size
plot(lines_HARV, add=TRUE,
     legend("bottom", NULL, labels, fill=labCols, bty="n", cex=.8))
```

<div id="challenge" markdown="1">
##Challenge: Specify Color By Attribute When Plotting

1. If it isn't a footpath, determine what TYPE the other two lines on the plot
are. 
2. Create a map to differentiate and determine what the other two paths are. 
3. Why might there be line types that appear in our shapefile but aren't
appearing on the plot? 
</div>
``` {r challenge-code-plot-other, results="hide", echo=FALSE}
#1 & 2 remind self what levels there are, then trial and error.  
levels(lines_HARV@data$TYPE)
# unk path in dark blue 
col[lineType == "boardwalk"] <- "darkblue"

#create base map w/ polygon & tower
plot(aoiBoundary_HARV, col = "lightgreen", 
     main="NEON Harvard Forest\nField Site")
plot(point_HARV, add  = TRUE, pch = 19, col = "purple")

#plot
plot(x=lines_HARV, col=col, add=TRUE)

#legend
labels= c("footpath", "boardwalk", "tower")
labCols=c("red", "darkblue", "purple")
plot(lines_HARV, add=TRUE,
     legend("bottom", NULL, labels, fill=labCols, bty="n", cex=.8))

#3. The extent of the lines is much larger than the field site boundary.  Since
# we plotted the field site boundary first only those lines in that extent are
# appearing on our plot.  
extent(aoiBoundary_HARV)
extent(lines_HARV)
```
