---
layout: post
title: "Shapefile attributes"
date:   2015-10-23 20:49:52
authors: "Joseph Stachelek, Leah Wasser"
dateCreated:  2015-10-23 20:49:52
lastModified: 2015-10-23 14:28:52
category: 
tags: []
mainTag: 
description: "This post explains the nature of shapefile attributes. Participants will be able to locate and query shapefile attributes as well as subset shapefiles by specific attribute values."
code1: 
image:
  feature: NEONCarpentryHeader_2.png
  credit: Graphic: National Ecological Observatory Network (NEON)
  creditlink: http://www.neoninc.org
permalink: /R/Raster-Data-In-R/
code1: /R/2015-07-22-Introduction-to-Raster-Data-In-R.R
comments:

---


##About
This post explains the nature of shapefile attributes. Participants will be able to locate and query shapefile attributes as well as subset shapefiles by specific attribute values.

###Goals / Objectives
After completing this activity, you will:

 * Be able to query shapefile attributes
 * Be able to subset shapefiles based on specific attributes
 
##Querying shapefile attributes

Recall from the previous lesson that shapefile metadata can include a _class_, a _features_ count, an _extent_, and a _coordinate reference system_ (crs). Shapefile _attributes_ include measurements that correspond to the geometry of the shapefile features.

Metadata can be extracted individually using the `class()`, `length()`, `extent()`, `crs()`, `R` commands. Attributes can be extracted using the `slot()` command. 

```{r Extract Individual Attributes}
#view class
class(x = squarePlot)
class(x = lines)
class(x = point)

#view features count
length(x = squarePlot)
length(x = lines)
length(x = point)

#view crs
crs(x = squarePlot)
crs(x = lines)
crs(x = point)

#view extent
extent(x = squarePlot)
extent(x = lines)
extent(x = point)

#view attributes
slot(object = squarePlot, name = "data")
slot(object = lines, name = "data")
slot(object = point, name = "data")
```


##Subsetting shapefiles

We already saw how to return all attributes of a spatial object using the `slot()` function. However, we can directly return individual attributes using the dollar sign notation. In the following code chunk we return a list of attribute names using the `slot()` and `names()` functions. Next, we query the _TYPE_ attribute individually. Finally, we subset the `lines` object using dollar sign notation. Note how our subsetting operation reduces the _features_ count from 13 to 2. 

```{r Subsetting shapefiles}
names(slot(object = lines, name = "data"))

lines$TYPE

lines[lines$TYPE == "footpath",]
```
