---
layout: post
title: "Lesson 02: CSV to Shapefile in R"
date:   2015-10-24
authors: [Joseph Stachelek, Leah Wasser]
contributors: [Megan A. Jones, Sarah Newman]
dateCreated:  2015-10-23
lastModified: `r format(Sys.time(), "%Y-%m-%d")`
category: 
tags: [module-1]
mainTag: GIS-Spatial-Data
description: "This lesson walks through how to convert a csv file contains coordinate information into a spatial object in R."
code1: 
image:
  feature: NEONCarpentryHeader_2.png
  credit: A collaboration between the National Ecological Observatory Network (NEON) and Data Carpentry
  creditlink: http://www.neoninc.org
permalink: /R/csv-to-shapefile-R/
comments: false
---

{% include _toc.html %}

#About

This lesson will review how to import spatial points stored in a  csv file into
R as a spatial object - a `SpatialPointsDataFrame`. We will also learn how to
reproject data imported in shapefile format into another projection, create a 
shapefile from a spatial object in R, and plot raster and vector data as overlays
in the same plot. 

**R Skill Level:** Intermediate - you've got the basics of `R` down.

<div id="objectives" markdown="1">

##Goals / Objectives
After completing this activity, you will know how to:

* Import csv files containing x,y coordinate locations into R.
* Convert a csv to a spatial object.
* Project coordinate locations provided in a Geographic Coordinate System
(Latitude, Longitude) to a projected coordinate system (UTM).
* Plot raster and vector data in the same plot to create a map.


###What you'll need

You will need the most current version of R or RStudio loaded on your computer 
to complete this lesson.

###R Libraries to Install:

* **raster:** `install.packages("raster")`
* **rgdal:** `install.packages("rgdal")`

<a href="{{ site.baseurl }}/R/Packages-In-R/" target="_blank"> 
More on Packages in R - Adapted from Software Carpentry.</a>

##Data to Download

Download the shapefiles needed to complete this lesson:

<a href="http://files.figshare.com/2387960/boundaryFiles.zip" class="btn btn-success"> 
DOWNLOAD Harvard Forest Shapefiles</a>

<a href="http://files.figshare.com/2387960/NEON_RemoteSensing.zip" class="btn btn-success"> 
DOWNLOAD Harvard Forest Remote Sensing</a>

###Recommended Reading
This lesson is a part of a series on vector and raster data in R.

1. <a href="{{ site.baseurl }}/R/open-shapefiles-in-R/">
Intro to shapefiles in R</a>
2. <a href="{{ site.baseurl }}/R/shapefile-attributes-in-R/">
Working With Shapefile Attributes in R </a>
3. <a href="{{ site.baseurl }}/R/csv-to-shapefile-R/">
CSV to Shapefile in R</a>
4. <a href="{{ site.baseurl }}/R/crop-extract-raster-data-R/">
Crop and extract raster values in R</a>

</div>

The `HARV_PlotLocations.csv` file contains x,y locations for some plots that 
NEON <a href="http://www.neoninc.org/science-design/collection-methods/terrestrial-organismal-sampling" target="_blank"> field crews are sampling vegetation and other 
ecological metrics. We would like to create a map that includes the locations of 
these plot locations. We also would like to export the data in a `shapefile` 
format to share with our colleagues. This shapefile could then be imported into
any GIS software.

Sometimes spatial data are downloaded or stored in a text file format (`.txt` or
`.csv`). We can convert these data in text format into a spatial object as an R 
`SpatialPointsDataFrame`. The `SpatialPointsDataFrame` format allows you to store
columns from your `data.frame` in an object so that you can bring along metadata
and attribute data. 
Note: There is a `SpatialPoints` object in R that does not allow you to store
associated `attributes`. 


```{r load-libraries }

#first load libraries
library(rgdal)
library(raster)

```

##Reading in a CSV
To begin, let's read in the `csv` file that contains the x,y coordinate locations---`HARV_PlotLocations.csv`.

```{r read-csv }

#set working directory to the directory location on your computer where
#you downloaded and unzipped the data files used in this lesson
#setwd("pathToDirHere")

#Read the csv file
plot.location <- read.csv("boundaryFiles/HARV/HARV_PlotLocations.csv")

#look at the data structure
head(plot.location)

```

#Challenge: Reading data structure
 * What is the coordinate reference system (CRS) for this data? 
 

When we create a spatial points dataframe from a .csv, there is no inherent
coordinate reference system (CRS) associated with it. We can assign a CRS if we
know what the coordinate system the x,y values are in.

In this case, the CRS for this data is UTM zone 18N (utmZone is 18N for all
points in the header we just looked at). Luckily this is the same as our boundary
shapefiles that we previously opened. 

If it is not still open in R, let's open up the shapefile `HarClip_UTMZ18`.

```{r read-shp }
#note: readGR is preferred as it maintains the projection infomation
aoiBoundary <- readOGR("boundaryFiles/HARV", "HarClip_UTMZ18")

```

This is the shapefile that we want to match up with our .csv file.  It is
important to first confirm that the CRS of our .csv coordinates is the same as 
the CRS of the shapefile. We can figure that out by looking at the metadata for
the shapefile.

```{r check-crs-shapefile-csv}
#check the CRS in shapefile
crs(aoiBoundary)
```

Next, we will convert our imported `plot.location` object to a
`SpatialPointsDataFrame` and assign it a CRS using the shapefile.

The x,y location values for the dataset are available and named "x" (column 1)
and "y" (column 2).  We can use that for our spatial data point objects. We know
the .csv file coordinates are in the same CRS as the other shapefiles. We can
assign this to the `SpatialPointsDataFrame` when we import it.

``` {r plot-csv-shapefile}
#plot the points
# Add title to map using main=""
plot.locationSp <- SpatialPointsDataFrame(plot.location[,1:2],    #the columns                                          #which represent x,y coordinate locations 
                    plot.location,    #The R object to convert to        
                                      #SpatialPointsDataframe
                    proj4string = crs(aoiBoundary))   # assinging a CRS system
                    #to our shapefile- in this instance we are pulling the CRS 
                    #from the aoiBoundary object
                                          
#look at CRS
crs(plot.locationSp)

```

Now, we have created a `SpatialPointsDataFrame` object in R from a csv file! 
The `R` object has a CRS that matches the CRS of our shapefiles. 

Let's plot our data!

```{r plot-data }
plot(plot.locationSp, main="Map of Study Area")

#let's add the polygon & tower point layer to our plot
plot(aoiBoundary, add=TRUE)
plot (point, col="red", add=TRUE)

#add roads to our plot, Lesson00: lines<-readOGR("boundaryFiles/HARV/", "HARV_roads")
plot(lines, add=TRUE)

```

What happenned to the tower?  The extent of these data covers a larger area than 
the aoiBoundary file that we originally originally projected.  

```{r extent-files}
extent(aoiBoundary)
extent (plot.locationSp)
```

#Create a New Shapefile from A Spatial Object

We can write an R spatial object to a shapefile using the `writeOGR` function 
in `rgdal`. To do this we need the following arguments

* the name of the spatial object (plot.locationSp)
* the directory where we want to save our shapefile
           (to use current = getwd() or you can specify a different path)
* the name of the new shapefile  (PlotLocations)
* the driver which specifies the file format (ESRI Shapefile)

We can now write the shapefile. 

``` {r write-shapefile}
#write a shapefile
writeOGR(plot.locationSp, getwd(), "PlotLocations", driver="ESRI Shapefile")

```

#Dealing with Spatial Objects in Different CRSs

Often, we have data that are in different CRSs. In order to map these objects in
in geographic space—so they line up on a map—we need to reproject one or the other.
We can use the `spTransform` function to reproject a vector.

In this case, we have a shapefile called `newPlots_latLon` that is in a 
geographic coordinate system (latitude and longitude) instead of the UTM
coordinate system of our other shapefiles. 

Let's read these data in and try to add them to our map without converting so we
can see what this looks like. 

```{r project-vectors }

#note: read ogr is preferred as it maintains prj info
newPlots <- readOGR("boundaryFiles/HARV", "newPlots_latLon")

#re-creating base plot; you may already have this
plot(plot.locationSp, main="Map of Study Area")
plot(aoiBoundary, add=TRUE)
plot(point, col="red", add=TRUE)
plot(lines, add=TRUE)

#add these points to our plot
plot(newPlots, add=TRUE, col=115, pch=19)

```

We added the newPlots to our existing plot and nothing happened?

What is the issue? Typically when data that we know should line up, doesn't,
there is a projection issue. Let's check it out.

First, let's look at the data in our new plots layer.Often the coordinates 
themselves can give us a clue as to whether the data are projected or are in 
latitude and longitude (geographic coordinate system).
We can use the `@coords` function to view the coordinates of the points in our
spatial object.

``` {r explore-spatial-data }
#view coordinates
newPlots@coords


#it appears as if our coordinates are in lat / long
#let's check the CRS of our data to see what's going on
crs(newPlots)

```

We've added points to our map. However, they are not appearing in our study area.
This is because our shapefile is in a geographic (latitude and longitude) reference 
system. They are not projected. We will need to project the data to `UTM` which is
what our other shapefiles are in to make things line up.

We will use `spTransform` to do project our data.


```{r project-data }

#let's double check the crs of our plots layer
crs(plot.locationSp)

#reproject our vector layer to the projection of our other plot data.
newPlots_UTMZ18N<- spTransform(newPlots,crs(plot.locationSp))


#plotting new projection
plot(newPlots_UTMZ18N, add=TRUE, col=116, pch=19)


```


#CHALLENGE: 


#Plotting - Order of Operations

Note that we had 2 points in our new plots layer, however the only one new point
appeared on our map. This is due to the extent of the map, which was defined by 
the order in which layers were added. 

What happens if we create a new plot, but add the roads layer first?

```{r plot-data-2 }

#add roads to our plot
plot(roads, main="Study Area")
#plot the points
#use main = to add a title to the map
plot(newPlots_UTMZ18N,add=T, col=116,pch=19)

plot(plot.locationSp, add=T)

#let's add the polygon layer to our plot
plot(aoiBoundary, add=TRUE)

```

#this could also be on your own -- they know how to plot rasters.

##Create a vector-raster overlay map

We will use the base graphics `plot()` command to create a vector-raster overlay 
plot. Next, we will load a raster file using the `raster` command, plot this 
raster, and overlay the lines and squareplot shapefiles using the `add` argument. We'll name the raster data `chm` as it is a Canopy Height Model.  

```{r Plot vector-raster overlay }

chm <- raster("NEON_RemoteSensing/HARV/CHM/HARV_chmCrop.tif")

#notice that the `\n` command allows you to split title lines into two.

plot(chm, main="Tree Height\nHarvard Forest")

plot(roads, add = TRUE)
plot(aoiBoundary, add = TRUE)
plot(plot.locationSp, add = TRUE, pch=19)

```

#Challenge - One Your Own

> to go here
>
>
