---
layout: post
title: "CSV to Spatial Points in R"
date:   2015-1-26 20:49:52
authors: "Joseph Stachelek, Leah Wasser"
dateCreated:  2014-11-26 20:49:52
lastModified: 2015-07-23 14:28:52
category: time-series-workshop
tags: [module-1]
mainTag: GIS-Spatial-Data
description: "About this post."
image:
  feature: 
  credit: 
  creditlink: http://www.neoninc.org
permalink: /R/csv-to-spatial-points/
code1: /R/2015-07-22-csv-to-spatial-points-In-R.R
comments: true
---

<section id="table-of-contents" class="toc">
  <header>
    <h3>Contents</h3>
  </header>
<div id="drawer" markdown="1">
*  Auto generated table of contents
{:toc}
</div>
</section><!-- /#table-of-contents -->



<div id="objectives">
<strong>R Skill Level:</strong> Beginner / Intermediate

<h3>Goals / Objectives</h3>
After completing this activity, you will:
<ol>
<li>Import a time series formated csv file.</li>
<li>Modify a date field that contains non-date characters and turn it into a proper
date formated field.</li>
<li>Plot the data.</li>
<li>Convert time to a local time zone.</li>
</ol>

<h3>R Libraries to Install</h3>
<ul>

</ul>

<h3>Data to Download</h3>
 

<h3>Pre-reqs</h3>
<p>.</p>
</div>

###Lesson 3 - How to convert csv with spatial locations to a spatial data frame


Objectives
* Understand how to convert a csv with spatial points to a spatial object in R.
* Convert to spatial points data frame
* Know how to plot a spatial points data
* Know how to create a basic map of a raster with a shapefile overlay
* Be able to compare and align the projections of raster and vector objects


Sometimes spatial data are downloaded or stored in a text file format (`txt` or csv`).
We can convert these data in text format, into a spatial object in R `SpatialPointsDataFrame`. Note that there is a `SpatialPoints` object in R that does not
allow you to store associated `attributes. The `SpatialPointsDataFrame allows you to store
columns from your `dataFrame` in the object.

To work with vector data in R, we can use the `rgdal` library. We will load the `raster`
library to work with rasters. 

```{r load-libraries }

#first load libraries
library(rgdal)
library(raster)

```

##Reading in our CSV

The `HARV_PlotLocations.csv` file contains x,y locations for some plots that 
NEON field crew are sampling vegetation and other ecological metrics. We would like
to create a map that includes the locations of these plot locations. We also would like
to export the data in a `shapefile` format to facilitate bring the data into a `GIS`.

To begin, let's read in the `csv` file that contains the x,y coordinate locations.

```{r read-csv }

#Leah's notes
setwd("~/Documents/data/1_DataPortal_Workshop/1_WorkshopData")

#set strings as factors to false
options(stringsAsFactors = FALSE)

#Read the csv file
plot.location <- read.csv("boundaryFiles/HARV/HARV_PlotLocations.csv")

#look at the data structure
head(plot.location)

```

When we create a spatial points dataframe from a csv, there is no inherent `CRS` associated
with it. We can assign a CRS if we know what the coordinate system the x,y values are in.

In this case, the CRS of our csv is UTM zone 18N (similar to our boundary shapefiles) that
we opened previously. If the object is not still open in R, let's open up a shapefile that is in that project. Next, we will convert our imported `plot.location` object to a `SpatialPointsDataFrame` and assign it a CRS using the shapefile.


NOTE: it is important to first CONFIRM that the CRS of our CSV coordinates is the 
SAME as the CRS of the shapefile. We can figure that out by looking at the metadata for
the CSV file which we hopefully have!

```{r}
#note: read ogr is preferred as it maintains prj info
squarePlot <- readOGR("boundaryFiles/HARV/","HarClip_UTMZ18")

#check the CRS
crs(squarePlot)

#the x,y location values for the dataset are availble in 
#the x and y columns. We can use that for our spatial data point object
#we know the csv file coordinates are in the same CRS as the other shapefiles.
#we can assign this to the spatial points data frame when we import it.
plot.location_spatial <- SpatialPointsDataFrame(plot.location[,1:2],
                                                plot.location,
                                                proj4string = crs(squarePlot))

#look at CRS
crs(plot.location_spatial)

```

Now, we have a `SpatialPointsDataFrame` object in R! The `R` object has a CRS that matches
the CRS of our shapefiles. Let's plot our data!


```{r}

#plot the points
#use main = to add a title to the map
plot(plot.location_spatial, main="Map of Study Area")

#let's add the polygon layer to our plot
plot(squarePlot, add=TRUE)

#add roads to our plot
roads <- readOGR("boundaryFiles/HARV/","HARV_roadStream")
plot(roads, add=TRUE)

```

#Create a New Shapefile from A Spatial Object

We can write an R spatial object to a shapefile using the `writeOGR` function 
in `rgdal`. To do this we need the following arguments

* the name of the spatial object
* the director where we want to save our shapefile
* the nae of the new shapefile
* the driver which specifies the file format (in this case a shapefile)


```{r }
#write a shapefile
writeOGR(plot.location_spatial, getwd(), "newFile", driver="ESRI Shapefile")

```

#Dealing with Spatial objects in different CRS'

Often, we have data that are in different CRS'. In order to map these objects
in the same visual space, we need to then reproject them. We can use the 
`spTransform` function to reproject a vector.

In this case, we have a shapefile called `newPlots_latLon` that is in a 
geographic coordinate system (latitude and longitude). Let's read these data in
and try to add them to our map

```{r}

#note: read ogr is preferred as it maintains prj info
newPlots <- readOGR("boundaryFiles/HARV/","newPlots_latLon")

#http://www.statmethods.net/advgraphs/parameters.html
#add these points to our plot
plot(newPlots, add=TRUE, col=115, pch=19)


#let's check the CRS of our data to see what's going on
crs(newPlots)


```

We've added points to our map. However, they are not appearing in our study area.
This is because our shapefile is in a GEOGRAPHIC (latitude and logiture) reference system.
We will need to reproject it to UTM to see if on our map. We will use `spTransform` to do
this.


```{r }

#let's double check the crs of our plots layer
crs(plot.location_spatial)

#reproject our vector layer to the projection of our other plot data.
newPlots_UTMZ18N<- spTransform(newPlots,crs(plot.location_spatial))
#try to plot the data
plot(newPlots_UTMZ18N, add=TRUE, col=116,pch=19)


```

##Create a vector-raster overlay map

We will use the base graphics `plot()` command to create a vector-raster overlay plot. 
Next, we will load a raster from disk user the `raster` command, plot this raster, and overlay the lines and squareplot shapefiles using the `add` argument. 

```{r Plot vector-raster overlay }

r <- raster("NEON_RemoteSensing/HARV/CHM/HARV_chmCrop.tif")

#notice that the `\n` command allows you to split title lines into two.

plot(r, main="Tree Height\nHarvard Forest")

plot(roads, add = TRUE)
plot(squarePlot, add = TRUE)

plot(plot.location_spatial, add = TRUE, pch=19)

```

#NEED TO ADD CHALLENGES!