---
layout: post
title: "Lesson 02: CSV to Shapefile in R"
date:   2015-10-24
authors: "Joseph Stachelek, Leah Wasser"
dateCreated:  2015-10-23
lastModified: 2015-10-26
tags: [module-1]
description: "This lesson walks through how to convert a csv file contains coordinate information into a spatial object in R."
image:
  feature: NEONCarpentryHeader_2.png
  credit: A collaboration between the National Ecological Observatory Network (NEON) and Data Carpentry
  creditlink: http://www.neoninc.org
permalink: /R/csv-to-shapefile-R/
---

<section id="table-of-contents" class="toc">
  <header>
    <h3>Contents</h3>
  </header>
<div id="drawer" markdown="1">
*  Auto generated table of contents
{:toc}
</div>
</section><!-- /#table-of-contents -->

#About
This lesson will review how to import spatial points, stored in a  csv file into
R as a spatial object - a `SpatialPointsDataFrame`. We will also learn how to
reproject data imported in shapefile format, into another projection, create a 
shapefile from a spatial object in R and plot raster and vector data as overlays
in the same plot. 

##Goals / Objectives
After completing this activity, you will:

* Import csv file containing x,y coordinate locations.
* Convert the csv to a spatial object
* Project coordinate locations provided in a Geographic Coordinate System (Latitude, Longitide) to projected (UTM)
* Plot raster and vector data in the same plot (create a map).


<div id="objectives">
<strong>R Skill Level:</strong> Beginner / Intermediate
<h3>What you'll need</h3>

You will need the most current version of R or R studio loaded on your computer 
to complete this lesson.

<h3>R Libraries to Install:</h3>
<ul>
<li><strong>raster:</strong> <code> install.packages("raster")</code></li>
<li><strong>rgdal:</strong> <code> install.packages("rgdal")</code></li>
<li><strong>ggplot2:</strong> <code> install.packages("ggplot2")</code></li>
</ul>

<a href="{{ site.baseurl }}/R/Packages-In-R/" target="_blank"> 
More on Packages in R - Adapted from Software Carpentry.</a>

<h2>Data to Download</h2>

Download the shapefiles neede to complete this lesson:

<a href="http://files.figshare.com/2387960/boundaryFiles.zip" class="btn btn-success"> 
DOWNLOAD Harvard Forest Shapefiles</a>

<h3>Recommended Reading</h3>
This lesson is a part of a series on vector and raster data in R.

<ol>
<li> <a href="{{ site.baseurl }}/R/open-shapefiles-in-R/">
Intro to shapefiles in R</a></li>
<li><a href="{{ site.baseurl }}/R/shapefile-attributes-in-R/">
Working With Shapefile Attributes in R </a></li>
<li><a href="{{ site.baseurl }}/R/csv-to-shapefile-R/">
CSV to Shapefile in R</a></li>
<li><a href="{{ site.baseurl }}/R/crop-extract-raster-data-R/">
Crop and extract raster values in R</a></li>
</ol>
</div>






Sometimes spatial data are downloaded or stored in a text file format (`txt` or 
csv`). We can convert these data in text format, into a spatial object in R `SpatialPointsDataFrame`. Note that there is a `SpatialPoints` object in R that 
does not allow you to store associated `attributes. The `SpatialPointsDataFrame 
allows you to store columns from your `data.frame` in the object.


```{r load-libraries }

#first load libraries
library(rgdal)
library(raster)

```

##Reading in our CSV

The `HARV_PlotLocations.csv` file contains x,y locations for some plots that 
(NEON)[http://www.neoninc.org] field crew are sampling vegetation and other 
ecological metrics. We would like to create a map that includes the locations of 
these plot locations. We also would like to export the data in a `shapefile` 
format to share with our colleagues. This shapefile can also be imported into any
`GIS` software.

To begin, let's read in the `csv` file that contains the x,y coordinate locations.

```{r read-csv }

#Leah's notes
setwd("~/Documents/data/1_DataPortal_Workshop/1_WorkshopData")

#Read the csv file
plot.location <- read.csv("boundaryFiles/HARV/HARV_PlotLocations.csv")

#look at the data structure
head(plot.location)

```

When we create a spatial points dataframe from a csv, there is no inherent `CRS` 
associated with it. We can assign a CRS if we know what the coordinate system 
the x,y values are in.

In this case, the CRS of our csv is UTM zone 18N (similar to our boundary shapefiles) 
that we opened previously. If the object is not still open in R, let's open up a 
shapefile that is in that project. Next, we will convert our imported `plot.location` 
object to a `SpatialPointsDataFrame` and assign it a CRS using the shapefile.


NOTE: it is important to first CONFIRM that the CRS of our CSV coordinates is the 
SAME as the CRS of the shapefile. We can figure that out by looking at the metadata 
for the CSV file which we hopefully have!

#NOTE: create metadata for this file!

```{r read-shp }
#note: read ogr is preferred as it maintains prj info
aoiBoundary <- readOGR("boundaryFiles/HARV/","HarClip_UTMZ18")

#check the CRS
crs(aoiBoundary)

#the x,y location values for the dataset are available in 
#the x and y columns. We can use that for our spatial data point object
#we know the csv file coordinates are in the same CRS as the other shapefiles.
#we can assign this to the SpatialPointsDataFrame when we import it.
plot.locationSp <- SpatialPointsDataFrame(plot.location[,1:2],
                                                plot.location,
                                                proj4string = crs(aoiBoundary))

#look at CRS
crs(plot.locationSp)

```

Now, we have created a `SpatialPointsDataFrame` object in R using a csv file! 
The `R` object has a CRS that matches the CRS of our shapefiles. Let's plot our 
data!


```{r plot-data }

#plot the points
#use main = to add a title to the map
plot(plot.locationSp, main="Map of Study Area")

#let's add the polygon layer to our plot
plot(aoiBoundary, add=TRUE)

#add roads to our plot
roads <- readOGR("boundaryFiles/HARV/","HARV_roadStream")
plot(roads, add=TRUE)

```

#Create a New Shapefile from A Spatial Object

We can write an R spatial object to a shapefile using the `writeOGR` function 
in `rgdal`. To do this we need the following arguments

* the name of the spatial object
* the director where we want to save our shapefile
* the name of the new shapefile
* the driver which specifies the file format (in this case a shapefile)


```{r write-shapefile }
#write a shapefile
writeOGR(plot.locationSp, getwd(), "newFile", driver="ESRI Shapefile")

```

#Dealing with Spatial objects in different CRS'

Often, we have data that are in different CRS'. In order to map these objects
in the same visual space, we need to then reproject them. We can use the 
`spTransform` function to reproject a vector.

In this case, we have a shapefile called `newPlots_latLon` that is in a 
geographic coordinate system (latitude and longitude). Let's read these data in
and try to add them to our map

```{r project-vectors }

#note: read ogr is preferred as it maintains prj info
newPlots <- readOGR("boundaryFiles/HARV/","newPlots_latLon")

#http://www.statmethods.net/advgraphs/parameters.html
#add these points to our plot
plot(newPlots, add=TRUE, col=115, pch=19)

```

We tried to add the newPlots to our existing plot and nothing happened? What
is the issue? Typically when data that we know should line up, doesn't, there
is a projection issue. Let's check it out.

First, let's look at the data in our new plots layer.
Often the coordinates themselves can give us a clue as to whether the data are
projected OR are in latitude and longitude (geographic coordinate system).
We can use the `@coords` function to view the coordinates of the points in our
spatial object.

``` {r explore-spatial-data }
#view coordinates
newPlots@coords

#it appears as if our coordinates are in lat / long
#let's check the CRS of our data to see what's going on
crs(newPlots)


```

We've added points to our map. However, they are not appearing in our study area.
This is because our shapefile is in a GEOGRAPHIC (latitude and longitude) reference 
system. They are not projected. We will need to project the data to `UTM` which is
what our other shapefiles are in to make things line up.

We will use `spTransform` to do project our data.


```{r project-data }

#let's double check the crs of our plots layer
crs(plot.locationSp)

#reproject our vector layer to the projection of our other plot data.
newPlots_UTMZ18N<- spTransform(newPlots,crs(plot.locationSp))
#try to plot the data
plot(newPlots_UTMZ18N, add=TRUE, col=116,pch=19)


```


#ON YOUR OWN -- 
#Plotting - Order of Operations

Note that we had 2 points in our new plots layer, however the only one new point
appeared on our map. This is due to the extent of the map, which was defined by 
the order in which layers were added. 

What happens if we create a new plot, but add the roads layer first?

```{r plot-data-2 }

#add roads to our plot
plot(roads, main="Study Area")
#plot the points
#use main = to add a title to the map
plot(newPlots_UTMZ18N,add=T, col=116,pch=19)

plot(plot.locationSp, add=T)

#let's add the polygon layer to our plot
plot(aoiBoundary, add=TRUE)

```

#this could also be on your own -- they know how to plot rasters.

##Create a vector-raster overlay map

We will use the base graphics `plot()` command to create a vector-raster overlay 
plot. Next, we will load a raster from disk user the `raster` command, plot this 
raster, and overlay the lines and squareplot shapefiles using the `add` argument. 

```{r Plot vector-raster overlay }

chm <- raster("NEON_RemoteSensing/HARV/CHM/HARV_chmCrop.tif")

#notice that the `\n` command allows you to split title lines into two.

plot(chm, main="Tree Height\nHarvard Forest")

plot(roads, add = TRUE)
plot(aoiBoundary, add = TRUE)
plot(plot.locationSp, add = TRUE, pch=19)

```

#Challenge - One Your Own

> to go here
>
>
