---
title: "Shapefile attributes"
author: "Joseph Stachelek"
date: "October 22, 2015"
output: html_document
---

##About
This post explains the nature of shapefile attributes. Participants will be able to locate and query shapefile attributes as well as subset shapefiles by specific attribute values.

###Goals / Objectives
After completing this activity, you will:

 * Be able to query shapefile attributes
 * Be able to subset shapefiles based on specific attributes
 
##Querying shapefile attributes

Recall from the previous lesson that shapefile attributes can include a _class_, a _features_ count, an _extent_, a _coordinate reference system_ (crs), and additional _attributes_.

These attributes can be extracted individually using the `class()`, `length()`, `extent()`, `crs()`, and `slot()` `R` commands. 

```{r Extract Individual Attributes}
#view class
class(squarePlot)
class(lines)
class(point)

#view features count
length(squarePlot)
length(lines)
length(point)

#view crs
crs(squarePlot)
crs(lines)
crs(point)

#view extent
extent(squarePlot)
extent(lines)
extent(point)

#view attributes
slot(squarePlot, "data")
slot(lines, "data")
slot(point, "data")
```


##Subsetting shapefiles

We already saw how to return all attributes of a spatial object using the `slot()` function. However, we can directly return individual attributes using the dollar sign notation. In the following code chunk we return a list of attribute names using the `slot()` and `names()` functions. Next, we query the _TYPE_ attribute individually. Finally, we subset the `lines` object using dollar sign notation. Note how our subsetting operation reduces the _features_ count from 13 to 2. 

```{r Subsetting shapefiles}
names(slot(lines, "data"))

lines$TYPE

lines[lines$TYPE == "footpath",]
```
