---
layout: post
title: "Lesson 02: When Vector Data Don't Line Up - Dealing with Projections & CRS
in R"
date:   2015-10-25
authors: [Joseph Stachelek, Leah Wasser, Megan A. Jones]
contributors: [Sarah Newman]
dateCreated:  2015-10-23
lastModified: `r format(Sys.time(), "%Y-%m-%d")`
packagesLibraries: [rgdal, raster]
category: 
mainTag: vector-data-workshop
tags: [vector-data, vector-data-workshop]
description: "This lesson will overview projections as handled by R and in concept."
code1: 02-csv-vector-raster-plotting.R
image:
  feature: NEONCarpentryHeader_2.png
  credit: A collaboration between the National Ecological Observatory Network (NEON) and Data Carpentry
  creditlink: http://www.neoninc.org
permalink: /R/vector-data-reproject-crs-R/
comments: false
---

{% include _toc.html %}

##About

In this tutorial, we will create a base map of our study site using a state and 
US country boundary accessed from the United States Census Bureau. We will learn 
how to manage vector data that don't line up. 

**R Skill Level:** Intermediate - you've got the basics of `R` down.

<div id="objectives" markdown="1">

#Goals / Objectives
After completing this activity, you will:

* Obj 2
* Obj 2
* Obj 2

##Things You’ll Need To Complete This Lesson
To complete this lesson: you will need the most current version of R, and 
preferably RStudio, loaded on your computer.

###Install R Packages

* **raster:** `install.packages("raster")`
* **rgdal:** `install.packages("rgdal")`
* **sp:** `install.packages("sp")`

* [More on Packages in R - Adapted from Software Carpentry.]({{site.baseurl}}R/Packages-In-R/)

##Data to Download
{% include/dataSubsets/_data_Site-Layout-Files.html %}

{% include/dataSubsets/_data_Airborne-Remote-Sensing.html %}

****

{% include/_greyBox-wd-rscript.html %}

**Vector Lesson Series:** This lesson is part of a lesson series on 
[vector data in R ]({{ site.baseurl }}self-paced-tutorials/spatial-vector-series). It is also
part of a larger 
[spatio-temporal Data Carpentry Workshop ]({{ site.baseurl }}self-paced-tutorials/spatio-temporal-workshop)
that includes working with
[raster data in R ]({{ site.baseurl }}self-paced-tutorials/spatial-raster-series) 
and  
[tabular time series in R ]({{ site.baseurl }}self-paced-tutorials/tabular-time-series).

</div>


Data Came from here

https://www.census.gov/geo/maps-data/data/cbf/cbf_state.html

##

 

<i class="fa fa-star"></i> **Data Tip:** more here
{: .notice}

We will use the `rgdal` and `raster` libraries in this tutorial. 

```{r load-libraries}

#load packages
library(rgdal)  #for vector work; sp package should always load with rgdal. 
library (raster)   #for metadata/attributes- vectors or rasters

#set working directory to data folder
#setwd("pathToDirHere")

```

##import US boundaries
note: we can also use the R map (maptools??) package which has many boundaries 
available for direct mapping. We will be using these layers for teaching 
purposes to understand how to deal with the CRS mismatch....

##Read US Boundary File


```{r read-csv }

#Read the .csv file
State.Boundary.US <- readOGR("NEON-DS-Site-Layout-Files/US-Boundary-Layers",
          "US-State-Boundaries-Census-2014")

#look at the data structure
class(State.Boundary.US)

```

Note: the Z-dimension error is normal, readOGR doesn't import z (vertical or height ) dimension
data by default.
http://www.inside-r.org/packages/cran/rgdal/docs/ogrInfo 

Next, let's plot the data 
```{r find-coordinates }

#view column names
plot(State.Boundary.US, main="Map of Continental US State Boundaries")

```

##US Boundary Layer  

Let's add another layer to out map - let's ouline the US in a bolder line


```{r check-out-coordinates }
#Read the .csv file
Country.Boundary.US <- readOGR("NEON-DS-Site-Layout-Files/US-Boundary-Layers",
          "US-Boundary-Dissolved-States")

#look at the data structure
class(Country.Boundary.US)

#view column names
plot(State.Boundary.US, 
     main="Map of Continental US State Boundaries",
     border="gray40")

#view column names
plot(Country.Boundary.US, 
     lwd=4, 
     border="gray18",
     add=TRUE)

```

Next, let's add a the tower location where our study area is located.


```{r explore-units}

#Import a point shapefile 
point_HARV <- readOGR("NEON-DS-Site-Layout-Files/HARV/",
                      "HARVtower_UTM18N")

#plot point - looks ok? 
plot(point_HARV, 
     pch = 19, 
     col = "purple",
     main="Harvard Fisher Tower Location")
```

``` {r layer-point-on-states }
#plot state boundaries  
plot(State.Boundary.US, 
     main="Map of Continental US State Boundaries",
     border="gray40")

#Add US border outline 
plot(Country.Boundary.US, 
     lwd=4, 
     border="gray18",
     add=TRUE)

#add point tower location
plot(point_HARV, 
     pch = 19, 
     col = "purple",
     add=TRUE)

```

Nother happened? Why?
Let's check out the CRS of both datasets to see if we can figure out what
is happening.


```{r crs-sleuthing} 

#view CRS of our site data
crs(point_HARV)

#view crs of census data
crs(State.Boundary.US)
crs(Country.Boundary.US)
```

It looks like our data are in different projections. We can tell this
but lookig at the `CRS` in `proj4` format

##Understanding CRS in Proj4 Format

The `CRS` for our data are given to us by `R` in `proj4` format. Let's break down
the pieces of `proj4` string. The string contains all of the individual `CRS` elements
that `R` or another `GIS` might need. Each element is specified with a `+` sign, 
similar to how a `.csv` file is delimited or broken up by a `,`. After each `+`
we see the `CRS` element being defined. For example `+proj=` and `+datum=`.


###UTM Proj4 String
Our project string below speficies the UTM projection as follows: 

`+proj=utm +zone=18 +datum=WGS84 +units=m +no_defs +ellps=WGS84 +towgs84=0,0,0` 

* **+proj=utm:** the projection is UTM, UTM has several zones.
* **+zone=18:** the zone is 18
* **datum=WGS84:** the datum WGS84 (the datum refers to the  0,0 reference for the 
coordinate system used in the projection)
* **+units=m:** the units for the coordinates are in METERS.
* **+ellps=WGS84:** the ellipsoid (how the earth's  roundness is calculated) for 
the data is WGS84

Note that the `zone` is unique to the UTM projection. Not all `CRS` will have a 
zone.

###Geographic (lat / long) Proj4 String
Our project string below speficies the UTM projection as follows: 

`+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0` 

* **+proj=longlat:** the data are in a geographic (latitute and longitude Coordinate System)
* **datum=WGS84:** the datum WGS84 (the datum refers to the  0,0 reference for the 
coordinate system used in the projection) 
* **+ellps=WGS84:** the ellipsoid (how the earth's  roundness is calculated) for 
the data is WGS84

Note that there are no specified units above. This is because this geographic 
coordinate reference system is typically in latitude and longitude which is most 
often recordedin `Decimal Degrees`.

<i class="fa fa-star"></i> **Data Tip:** +towgs84=0,0,0  conversion from current datum/ellipse to WGS84 -- nothing,
we're already in WGS84 for both. 
{: .notice}

#view object Extent

TO further understand the uhits that our data are in, let's view the extent for
each object.

```{r view-extent }

#extent for HARV in UTM
extent(point_HARV)

#extent for object in geographic
extent(State.Boundary.US)


```

Note the difference in the units for each object. The extent for `State.Boundary.US`
is in latitude and longitude. Whereas our data in UTM, is represented in meters.


***
##Proj4 & CRS Resources

* For more information on <a href="http://proj.maptools.org/faq.html" target="_blank"> proj4.</a>
* For information on datums in `R`: `projInfo(type = "datum")`
***


#Reproject Vector Data

Now we know our data in different `CRS`. To address this, we have to  modify or
 `reproject` our data so they are all in the *same* `CRS`. We can use `spTransform`
 to do this.When we make the conversion, we also have to specify 
the `crs` that we wish to transform our data to. As we saw above, this `CRS` contains
the `datum`, units and other information that `R` needs to `reproject` our data.
The `spTransform` function requires two inputs:

1. the name of the object that you wish to transform
2. the `CRS` that you wish to transform that object too. In this case we can 
use the `CRS` of the `State.Boundary.US` object as follows:

`crs(State.Boundary.US)`

NOTE: spTransform will only work if your original spatial object has a `CRS` assigned 
to it AND if that `CRS` is the correct `CRS`!

Next, let's reproject our point layer into the geographic - latitude and longitude 
`WGS84` coordinate reference system (CRS).


```{r crs-sptranform } 

#reproject data
point_HARV_WGS84 <- spTransform(point_HARV,
                                crs(State.Boundary.US))

#what is the CRS of the new object
crs(point_HARV_WGS84)
#does the extent look like decimal degrees?
extent(point_HARV_WGS84)
```

```{r plot-again }

#plot state boundaries  
plot(State.Boundary.US, 
     main="Map of Continental US State Boundaries\n With Fisher Tower Location",
     border="gray40")

#Add US border outline 
plot(Country.Boundary.US, 
     lwd=4, 
     border="gray18",
     add=TRUE)

#add point tower location
plot(point_HARV_WGS84, 
     pch = 19, 
     col = "purple",
     add=TRUE)

```


Reprojecting our data ensured that things line up on our map!


<div id="challenge" markdown="1">
##Challenge - Reprojecting Data

Create a map of the North Eastern United States as follows:

1. Import and plot `Boundary-US-State-NEast.shp`. Adjust line width as necessary.
2. REPROJECT the layer into UTM zone 18 north.
3. Layer the Fisher Tower point location `point_HARV` on top of the above plot.
4. Add a title to your plot.
5. Add a LEGEND to your plot that shows both the state boundary (line) and the 
Tower location POINT.  


</div>

```{r challenge-code-MASS-Map,  include=TRUE, results="hide", echo=FALSE, warning=FALSE}
#import mass boundary layer
#Read the .csv file
NE.States.Boundary.US <- readOGR("NEON-DS-Site-Layout-Files/US-Boundary-Layers",
          "Boundary-US-State-NEast")
#view crs
crs(NE.States.Boundary.US)

#create CRS object
UTM_CRS <- crs(point_HARV)
UTM_CRS

#reproject line and point data
NE.States.Boundary.US.UTM  <- spTransform(NE.States.Boundary.US,
                                UTM_CRS)
NE.States.Boundary.US.UTM

#plot state boundaries  
plot(NE.States.Boundary.US.UTM , 
     main="Map of North Eastern US\n With Fisher Tower Location - UTM Zone 18n",
     border="gray18",
     lwd=2)

#add point tower location
plot(point_HARV, 
     pch = 19, 
     col = "purple",
     add=TRUE)

#add legend
#to create a custom legend, we need to fake it
legend("bottomright", 
       legend=c("State Boundary","Fisher Tower"),
       lty=c(1,NA),
       pch=c(NA,19),
       col=c("gray18","purple"),
       bty="n")

```
