---
layout: post
title: "Lesson 03: Convert from .csv to Shapefile in R"
date:   2015-10-24
authors: [Joseph Stachelek, Leah Wasser, Megan A. Jones]
contributors: [Sarah Newman]
dateCreated:  2015-10-23
lastModified: `r format(Sys.time(), "%Y-%m-%d")`
packagesLibraries: [rgdal, raster]
category: 
mainTag: vector-data-workshop
tags: [vector-data, vector-data-workshop]
description: "This lesson walks through how to convert a .csv file that contains
coordinate information into a spatial object in R."
code1: 02-csv-vector-raster-plotting.R
image:
  feature: NEONCarpentryHeader_2.png
  credit: A collaboration between the National Ecological Observatory Network (NEON) and Data Carpentry
  creditlink: http://www.neoninc.org
permalink: /R/csv-to-shapefile-R/
comments: false
---

{% include _toc.html %}

##About

This lesson will review how to import spatial points stored in `.csv` (Comma
Separated Value) format into
`R` as a spatial object - a `SpatialPointsDataFrame`. We will also
reproject data imported in a shapefile format, and export a shapefile from an
`R` spatial object and plot raster and vector data as
layers in the same plot. 

**R Skill Level:** Intermediate - you've got the basics of `R` down.

<div id="objectives" markdown="1">

#Goals / Objectives
After completing this activity, you will:

* Be able to import .csv files containing x,y coordinate locations into `R`.
* Know how to convert a .csv to a spatial object.
* Understand how to project coordinate locations provided in a Geographic
Coordinate System (Latitude, Longitude) to a projected coordinate system (UTM).
* Be able to plot raster and vector data in the same plot to create a map.

##Things You’ll Need To Complete This Lesson
To complete this lesson: you will need the most current version of R, and 
preferably RStudio, loaded on your computer.

###Install R Packages

* **raster:** `install.packages("raster")`
* **rgdal:** `install.packages("rgdal")`
* **sp:** `install.packages("sp")`

* [More on Packages in R - Adapted from Software Carpentry.]({{site.baseurl}}R/Packages-In-R/)

##Data to Download
{% include/dataSubsets/_data_Site-Layout-Files.html %}

{% include/dataSubsets/_data_Airborne-Remote-Sensing.html %}

****

{% include/_greyBox-wd-rscript.html %}

**Vector Lesson Series:** This lesson is part of a lesson series on 
[vector data in R ]({{ site.baseurl }}self-paced-tutorials/spatial-vector-series). It is also
part of a larger 
[spatio-temporal Data Carpentry Workshop ]({{ site.baseurl }}self-paced-tutorials/spatio-temporal-workshop)
that includes working with
[raster data in R ]({{ site.baseurl }}self-paced-tutorials/spatial-raster-series) 
and  
[tabular time series in R ]({{ site.baseurl }}self-paced-tutorials/tabular-time-series).

</div>

##Spatial Data in Text Format

The `HARV_PlotLocations.csv` file contains `x, y` (point) locations for plot locations 
<a href="http://www.neoninc.org/science-design/collection-methods/terrestrial-organismal-sampling" target="_blank">  
where NEON </a>
samples vegetation and other ecological metrics. We would like to:

* Create a map of these plot locations. 
* Export the data in a `shapefile` format to share with our colleagues. This
shapefile can be imported into any GIS software.
* Create a map showing vegetation height with plot locations layered on top.

Spatial data are sometimes stored in a text file format (`.txt` or `.csv`). If 
the text file has an associated `x` and `y` location column, then we can 
convert it into an `R` spatial object which in the case of point data,
will be a `SpatialPointsDataFrame`. The `SpatialPointsDataFrame` 
allows us to store both the `x,y` values that represent the coordinate location
of each point and the associated attribute data - or columns describing each
feature in the spatial object.  

<i class="fa fa-star"></i> **Data Tip:** There is a `SpatialPoints` object (not
`SpatialPointsDataFrame`) in `R` that does not allow you to store associated
attributes. 
{: .notice}

We will use the `rgdal` and `raster` libraries in this tutorial. 

```{r load-libraries}

#load packages
library(rgdal)  #for vector work; sp package should always load with rgdal. 
library (raster)   #for metadata/attributes- vectors or rasters

#set working directory to data folder
#setwd("pathToDirHere")

```


##Import .csv 
To begin we will read in the .csv file that contains the `x, y` coordinate
plot locations at the NEON Harvard Forest Field Site (`HARV_PlotLocations.csv`).
Note we set `stringsAsFactors=FALSE` so our data import as a
`character` rather than a `factor` class.

```{r read-csv }

#Read the .csv file
plot.locations_HARV <- 
  read.csv("NEON-DS-Site-Layout-Files/HARV/HARV_PlotLocations.csv",
           stringsAsFactors = FALSE)

#look at the data structure
str(plot.locations_HARV)

```

Also note that `plot.locations_HARV` is a `data.frame` that contains 21 
locations (rows) and 15 variables (attributes). 

The next thing that we need to do is identify the columns containing coordinate
values. If we are lucky, our `.csv` will contain either:

* columns labeled "X" and "Y" OR
* Latitude and Longitude OR
* easting and northing (UTM coordinates)

Let's check out the column `names` of our file to see if we have coordinate locations.

```{r find-coordinates }

#view column names
names(plot.locations_HARV)

```

##Identify X,Y Location Columns

Note that our object is not yet a spatial object. However it does contain several 
fields of interest to us. The `plot.locations_HARV$easting` and 
`plot.locations_HARV$northing` columns contain coordinate location. 

```{r check-out-coordinates }
#view first 6 rows of the X and Y columns
head(plot.locations_HARV$easting)
head(plot.locations_HARV$northing)

#Note that  you can also call the same two columns using their COLUMN NUMBER
#view first 6 rows of the X and Y columns
head(plot.locations_HARV[,1])
head(plot.locations_HARV[,2])

```

In order to convert our `data.frame` to a `SpatialPointsDataFrame`, we need to know
the `CRS`. 

There are several ways to figure out the CRS of spatial data in text format.  

1. We can check the file `metadata` in hopes that the `CRS` was recorded in the data.
For more information on metadata, check out the
[Understanding Time Series Metadata]({{site.baseurl}}/R/Time-Series-Metadata) 
lesson. 
2. We can explore the file itself to see if `CRS` information is embedded in the 
file header or somewhere in the data.


Following the `easting` and `northing` columns, there is a `geodeticDa` and a 
`utmZone` column. These appear to contain `CRS` information (`datum` and `projection`).
Let's view those next. 


```{r view-CRS-info }
#view first 6 rows of the X and Y columns
head(plot.locations_HARV$geodeticDa)
head(plot.locations_HARV$utmZone)

```

It is not typical to store `CRS` information in a column. But this particular file
contains `CRS` information this way.

Viewing our data, we can see that the following fields appear to be spatial 
attributes: 

* `geodeticDa`: WGS84  -- this is geodetic datum WGS84
* `utmZone`: 18N


In 
[Lesson 02 - When vector data don't line up - CRS]({{site.baseurl}}/R/vector-data-reproject-crs-R/)
we learned about the components of a `proj4` string. We have everything we need 
to now assign a `CRS` to our data.frame.

We could create a `proj4` string, but if we have other data in the `UTM Zone 18N`
projection, it's much easier to simply copy the string directly! Let's import
the roads layer from Harvard forest and check out its `CRS`.

```{r explore-units}

#Import the line shapefile
lines_HARV <- readOGR( "NEON-DS-Site-Layout-Files/HARV/", "HARV_roads")

#view CRS
crs(lines_HARV)

#view extent
extent(lines_HARV)


```

Exploring the data above, we can see that the lines shapefile is in `UTM zone 18N`.
We can thus use the CRS from that spatial object to convert our non-spatial
`data.frame` into a `spatialPointsDataFrame`. 

Next, let's create a `crs` object that we can use to define the `CRS` of our 
`SpatialPointsDataFrame` when we create it

```{r crs-object } 
#create crs object
utm18nCRS <- crs(lines_HARV)
utm18nCRS

class(utm18nCRS)
```



##.csv to R SpatialPointsDataFrame

Next, let's convert our `data.frame` into a `SpatialPointsDataFrame`. To do this,
we need to specify:

1. The columns containing X (`easting`) and Y (`northing`) coordinate values
2. The CRS that the column coordinate represent (units are included in the CRS) -
stored in our `utmCRS` object.
3. optional: the other columns stored in the data frame that you wish to append
as attributes to your spatial object

We will use the `SpatialPointsDataFrame()` function to perform the conversion.

```{r convert-csv-shapefile}
#note that the easting and northing columns are in columns 1 and 2
plot.locationsSp_HARV <- SpatialPointsDataFrame(plot.locations_HARV[,1:2],
                    plot.locations_HARV,    #the R object to convert
                    proj4string = utm18nCRS)   # assign a CRS 
                                          
#look at CRS
crs(plot.locationsSp_HARV)

```

## Plot Spatial Object 
We now have a spatial `R` object, we can plot our newly created spatial object.

```{r plot-data-points }
plot(plot.locationsSp_HARV, 
     main="Map of Plot Locations")

# Save plot 
Harv_plotLocMap<- recordPlot()
```

While our points do plot the map isn't very informative.  We can bring in other
spatial objects to create a better map. 

Let's bring in shapefiles for a field site boundary, roads & trails, and a tower
location in Harvard Forest. (If you completed
[Shapefile Metadata & Attributes in R]({{site.baseurl}}/R/shapefile-attributes-in-R/)
lesson you already have these `R` objects).

```{r read-shp }
#note: readOGR is preferred as it maintains the projection infomation
#Import Field site boundary
aoiBoundary_HARV <- readOGR("NEON-DS-Site-Layout-Files/HARV",
                            "HarClip_UTMZ18")

#Import a roads/trails
lines_HARV <- readOGR( "NEON-DS-Site-Layout-Files/HARV/", "HARV_roads")

#Import a tower point
point_HARV <- readOGR("NEON-DS-Site-Layout-Files/HARV/",
                      "HARVtower_UTM18N")

```

These are shapefiles that we want to match up with our, now converted, .csv
file.  It is important to first confirm that both spatial objects have the same 
CRS.

```{r check-crs-shapefile-csv}
#check the CRS in shapefile
crs(plot.locationsSp_HARV)
crs(aoiBoundary_HARV)
crs(lines_HARV)
crs(point_HARV)
```

Yep, they are all the same.  We can now plot the shapefiles together.  

``` {r plot-data-maps}
#recall the original plot location map; optional
Harv_plotLocMap

#Add Field Site Boundary polygon & tower point layer to our "Map of Study Area"
plot(aoiBoundary_HARV, add=TRUE)
plot (point_HARV, col="red", add=TRUE)
plot(lines_HARV, add=TRUE)

# save this newer map
Harv_plotLocMap2<- recordPlot()
```

What happened to the tower?

The extent of the study plot locations covers a larger area than the
`aoiBoundary_HARV` file that we originally originally projected.  

```{r extent-files}
extent(aoiBoundary_HARV)
extent (plot.locationsSp_HARV)
```

#Borrow a CRS When Converting to a Spatial.Data.Frame
When we just convert our imported `plot.location_HARV` object to a
`SpatialPointsDataFrame` we assign it a `CRS` using a written out `proj4string`.
However, if we'd already had a Spatial.Data.Frame object with the same `CRS`)
(such as `aoiBoundary_HARV`, `lines_HARV` or `point_HARV`) we could have simply
borrowed the `CRS` from that object to use in the conversion.  

We'd just identify the file from which we'd use the `CRS` when doing the
conversion.  

``` {r plot-csv-shapefile}

plot.locations_HARV_matchCRS <-
               SpatialPointsDataFrame(plot.locations_HARV[,23:24],
                                      plot.locations_HARV,    # object to convert
                                      proj4string = crs(aoiBoundary_HARV))
                                    #assign CRS to match `aoiBoundary_HARV CRS
#check to make sure it worked
crs(plot.locations_HARV_matchCRS)
```

WARNING: We still have to know the correct `CRS` for the data and 
use a file with a matching `CRS`.  This doesn't solve our problems if we DO NOT 
know our `CRS`. 

#Create a New Shapefile from A Spatial Object

We can write an `R` spatial object to a shapefile using the `writeOGR` function 
in `rgdal`. To do this we need the following arguments:

* the name of the spatial object (`plot.locationsSp_HARV`)
* the directory where we want to save our shapefile
           (to use current = getwd() or you can specify a different path)
* the name of the new shapefile  ("PlotLocations_HARV")
* the driver which specifies the file format (ESRI Shapefile)

We can now write the shapefile. 

``` {r write-shapefile}
#write a shapefile
writeOGR(plot.locationsSp_HARV, getwd(),
         "PlotLocations_HARV", driver="ESRI Shapefile")

```


#Spatial Objects in Different CRSs
Too often, we have data that are in different `CRS`s. In order to map these
objects in geographic space we need to reproject one or the other so that 
they will line up on a map.

Let's explore reprojecting a spatial data object using a new study plot 
location (shapefile: "newPlot_latLon").

```{r project-vectors }
#note: readOGR is preferred as it maintains prj info
newPlots_HARV <- readOGR("NEON-DS-Site-Layout-Files/HARV",
                         "newPlots_latLon")

#add these points to our existing plot  "Map of Study Area"
Harv_plotLocMap2
plot(newPlots_HARV, add=TRUE, col=115, pch=19)

#save new plot
Harv_plotLocMap3<- recordPlot()
```

We added the new plot locations to our existing plot and nothing happened.  Why?

What is the issue? Typically when data that we know should line up doesn't,
there is a projection issue. (Plus we should ALWAYS check projections/ CRS prior
to starting to work with any new spatial data.)

Often the coordinates can give us a clue as to whether the data are projected or
are in latitude and longitude (a geographic coordinate system).
We can use the `@coords` function to view the coordinates of the points in our
spatial object.

``` {r see-coords }
#view coordinates
newPlots_HARV@coords
```

It appears as if our coordinates for the two new plot locations are in latitude
and longitude. Let's check the CRS of our data to see if it is also in latitude
and longitude.
``` {r see-crs}
crs(newPlots_HARV)
```

Our `newPlots_HARV` shapefile is in a geographic coordinate system (latitude and 
longitude) instead of the UTM coordinate system of our other shapefiles. These
coordinates are not projected, and we need to reproject them in order to make
things line up in our map.

We can use the `spTransform` function to reproject a vector.

```{r project-data }
#double check the CRS of our plots layer
crs(plot.locationsSp_HARV)

#reproject our vector layer to the projection of our other plot data.
newPlots_HARV_UTM<- spTransform(newPlots_HARV,
                               crs(plot.locationsSp_HARV))

#check crs
crs(newPlots_HARV_UTM)
```

Now that the `CRSs` match we can add our new plot points to our existing map.  

```{r plot-reprojected-points}
#plot new plot location -- reprojected
#col & pch parameters to make it stand out!
Harv_plotLocMap3
plot(newPlots_HARV_UTM, add=TRUE, col=116, pch=19) 

```

Yes!  A nice blue point showed up for our new plot... but wait only one.  We 
have two new plots.  Why did only one appear?  

#Plotting - Order of Operations
We had 2 points in our new plots layer, however the only one new point
appeared on our map. This is due to the extent of the map, which is defined by 
the extent of the first layer plotted (all others are "added" to that layer). 

What happens if we create a new plot, but add the roads layer first?

```{r plot-data-order }
#add roads to our plot
plot(lines_HARV, main="Study Area")

#add the study plot locations 
plot(newPlots_HARV_UTM,add=T, col=116)
plot(plot.locationsSp_HARV, add=T)

#add the boundary layer 
plot(aoiBoundary_HARV, col="red", add=TRUE)

#add the tower 
plot(point_HARV, add =TRUE, pch=16)

```

We can see all the study plot locations now because we plotted the roads/trails 
layer (`lines_HARV`) first and it is the layer with the largest extent.  

#Plotting - Vector & Raster Layers Together
To create a visually pleasing map or to combine all the data we are interested
in we often need to combine raster and vector layers into one figure. We can
use the base `R` `plot()` command to create a vector-raster layered 
plot. 

We will use a Canopy Height Model (`chm`) from the Harvard forest as the base
layer for a map of our study locations.  

```{r Plot vector-raster overlay }

chm_HARV <- raster("NEON-DS-Airborne-RemoteSensing/HARV/CHM/HARV_chmCrop.tif")

plot(chm_HARV,
     main="Map of Study Plots\n w/ Canopy Height Model\nNEON Harvard Forest")

plot(lines_HARV, add = TRUE)
plot(aoiBoundary_HARV, border="forestgreen", add = TRUE)
plot(plot.locationsSp_HARV, add = TRUE)
plot(newPlots_HARV_UTM, col="blue", add=TRUE)
plot(point_HARV, pch=17, add=TRUE)

```

<div id="challenge" markdown="1">
##Challenge: Useful Study Area Maps
Using a Digital Terrain Model
("NEON-DS-Airborne-Remote-Sensing/HARV/DTM/HARV_dtmCrop.tif") 
create a usable map to the study area.  Include the following: 

1. Useful title, figure legends, and axes labels. 
2. Any or all vector layers that we have created. 
3. A customized color ramp. 
4. A color-coded system to identify the different types of roads or trails that
occur in the study site.  

For assistance consider using the 
[Shapefile Metadata & Attributes in R]({{site.baseurl}}/R/shapefile-attributes-in-R/),   
the [Plot Raster Data in R]({{site.baseurl}}/R/Plot-Rasters-In-R/ )   
or the
[Plot Raster Time Series Data in R Using RasterVis and Levelplot]({{site.baseurl}}/R/Plot-Raster-Times-Series-Data-In-R/) 
lessons. 
</div>

```{r challenge-code-plotting-pretty-map,  include=TRUE, results="hide", echo=FALSE}

#load DTM
DTM_HARV <- raster("NEON-DS-Airborne-RemoteSensing/HARV/DTM/HARV_dtmCrop.tif")

#create basic plot
plot(DTM_HARV,
     main="Map of Study Plots\n w/ Digital Terrain Model\nNEON Harvard Forest")

plot(aoiBoundary_HARV, border="forestgreen", add = TRUE)
plot(plot.locationsSp_HARV, pch=20, add = TRUE)
plot(newPlots_HARV_UTM, pch=20, col="blue", add=TRUE)
plot(point_HARV, pch=17, add=TRUE)

#legend
labels= c("footpath", "boardwalk", "tower")
labCols=c("red", "darkblue", "black")

#add lines w/ legend & different colors
#remind self what levels there are 
levels(lines_HARV@data$TYPE)

#create linetype 
lineType <- lines_HARV@data$TYPE

#set color for all lines not later specified
col <- rep(x= "black", length = length(lineType))

# color the boardwalks
col[lineType == "boardwalk"] <- "darkblue"
plot(lines_HARV, add=TRUE,
     legend("bottom", NULL, labels, fill=labCols, bty="n", cex=.8))
```
