---
title: "Basic raster-vector integration"
author: "Joseph Stachelek"
date: "October 22, 2015"
output: html_document
---

##About

This post explains how to visualize raster and vector objects in the same plotting window.

###Goals / Objectives
After completing this activity, you will:

 * Know how to create a basic map of a raster with a shapefile overlay
 * Be able to compare and align the projections of raster and vector objects
 
##Create a vector-raster overlay map

We will use the base graphics `plot()` command to create a vector-raster overlay plot. The following code chunk loads a raster from disk user the `raster` command, plot this raster, and overlays the lines and squareplot shapefiles using the `add` argument. 

```{r Plot vector-raster overlay}
r <- raster(file.path(prepath, "data", "NEON_RemoteSensing", "HARV", "CHM", "HARV_chmCrop.tif"))

plot(r)
plot(lines, add = TRUE)
plot(squarePlot, add = TRUE)
```

##Compare and align vector/raster projections

Recall from lesson 2 that we can query the _coordinate reference system_ of a spatial object using the `crs()` function.

```{r }
(r_crs <- crs(r))
(squarePlot_crs <- crs(squarePlot))
(lines_crs <- crs(lines))

```

```{r }
spTransform(lines, )
projectRaster(r, )
```