---
title: "Advanced raster-vector integration"
author: "Joseph Stachelek"
date: "October 22, 2015"
output: html_document
---

##About

This post explains how to modify (crop) a raster extent based on the extent of a vector shapefile. Particpants will also be able to extract values from a raster that correspond to the geometry of a vector overlay.

###Goals / Objectives
After completing this activity, you will:

 * Be able to crop a raster to the extent of a vector layer
 * Be able to extract values from raster that correspond to a vector overlay
 
##Crop a raster to vector layer extent

The following code chunk crops our previously loaded raster by the extent of our polygon shapefile.

```{r Crop by vector extent}
r_cropped <- crop(x = r, y = squarePlot)
```

We can also use an `extent` object as input to the `y` argument to `crop()`. The `drawExtent()` function is an easy (but imprecise) way to construct an `extent` object. See the documentation for the `extent()` function for more ways to create an `extent` object (`help.search("extent", package = "raster")). 

```{r Crop by drawn extent}
extent <- raster::drawExtent()
r_cropped_man <- crop(x = r, y = extent)

```

##Extract values from a raster that correspond to a vector overlay

We can extract values from our raster layer that correspond to our vector shapefiles by using the `extract()` function. The basic call to `extent()` simply returns all raster values that are _touched_ by the shapefile components. More complicated calls to `extract()` might involve pulling points that fall within a specified `buffer` radius around our shapefile geometries. See the documentation for the `extract()` function for more details (`help.search("extract", package = "raster")`). Note that the `lines_extract` and `poly_extract` objects are of class `list`. Each component of this list correspondes to a discrete feature of the vector layer. Our call to the `summary()` function uses `unlist()` to collapse the raster values for these features.

```{r Extract from raster}
lines_extract <- extract(r, lines)
point_extract <- extract(r, point)
poly_extract <- extract(r, squarePlot)

class(lines_extract)
class(point_extract)
class(poly_extract)

summary(unlist(lines_extract))
summary(point_extract)
summary(unlist(poly_extract))

```